Microlensing as a possible probe of event-horizon structure in quasars

### 2017-06-27 paper1/ms.tex
editing

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, is a phenomenon seen in quasars
that have been multiply imaged by galaxies.  Individual images in a
multiple-image system can undergo sharp and seemingly random
brightness changes.  This can occur, even for constant intrinic
luminosity, as a consequence of two things: the very small size of the
central engine, and granularity of the mass distribution of a lensing
galaxy due to stars.  The latter means that the local magnification is
not a smooth function of source position. It contains a complicated
network of singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing has been first observed in the optical
range \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
X-rays \citep[e.g.,][]{2016arXiv160207601N}.  In principle, it exists
at all wavelengths \citep[and has been argued for at cm wavelengths
  by][]{2000A&A...358..793K} but will be washed out if the light is
from larger regions than the central engine of the quasar.  The effect
thus supplies an upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources, and then fit these mock lightcurves to source models
using Markov-chain Monte Carlo (MCMC).

@@

We now proceed to fit the mock lightcurve.  The basic fitting
technique is the same as used in several previous studies
\citep[e.g.,][]{2010ApJ...712..658P}.  Markov-Chain Monte-Carlo is
used to fit source parameters and additional nuisance parameters (see
below) jointly, and then marginalise out all uninteresting parameters,
leaving posterior distributions for any chosen parameter.

In addition to the source parameters, the fitting procedure includes
three nuisance parameters, namely the brightness normalisation and the
start and end points on the path.  The latter two are equivalent to
considering the location of the caustic with respect to the starting
point and the velocity component of the source transverse to the
caustic (that is, $v_{\!\perp}$).  The orientation of the source track
is another nuisance parameter, which we do not include it separately.
This is because, for a pure fold, changing the orientation of the
source track is equivalent to changing $v_{\!\perp}$ and the
orientation of the source (see Figure~\ref{fig:infinite_fold}), all
parameters that are included.  In the present case, the caustic is a
clean but non-ideal fold, hence disregarding the orientation of the
source track is expected to introduce some systematic error.

Figure \ref{fig:cc_forward} shows a fit to the mock lightcurve and
recovery of $r_{1/2}$ and $R_n/R_p$.

Figure \ref{fig:cc_backward} shows the result when a different caustic
is used to fit, than was used to generate the mock lightcurve.  This
mimics the unavoidable systematic error of not knowing the caustic
exactly.  The fit is slightly degraded, but still acceptable.

Figure \ref{fig:gc_forward} shows the result of attempting to fit a
Gaussian source.  There is a small but statistically very significant
misfit because a Gaussian curve cannot reproduce the part of the
lightcurve with the slight flattening and then a second sub-peak
(cf.~Figure~\ref{fig:char_points}) as the dark part of the crescent crosses the caustic.  
A hint of
such a feature is present in recent caustic-crossing observations of
the well-known system Q2237+0305 \citep[see Figure~3
in][]{2015ApJ...814L..26M} though interpretation as a crescent source
is speculation at present.

>>>>

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, is a phenomenon seen in quasars
that have been multiply imaged by galaxies.  Individual images in a
multiple-image system can undergo sharp and seemingly random
brightness changes.  This can occur, even for constant intrinic
luminosity, as a consequence of two things: the very small size of the
central engine, and granularity of the mass distribution of a lensing
galaxy due to stars.  The latter means that the local magnification is
not a smooth function of source position. It contains a complicated
network of singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is observable at optical
wavelengths \citep[e.g.,][]{2012A&A...544A..62S} and in
X-rays \citep[e.g.,][]{2016arXiv160207601N}.  In principle, it exists
at all wavelengths \citep[and has been argued for at cm wavelengths
  by][]{2000A&A...358..793K} but will be washed out if the light is
from larger regions than the central engine of the quasar.  The effect
thus supplies an upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources, and then fit these mock lightcurves to source models
using Markov-chain Monte Carlo (MCMC).  The general strategy is the same
as in previous studies \citep[e.g.,][]{2010ApJ...712..658P} though
the details are different.

@@

We now proceed to fit the mock lightcurve.  The basic fitting
technique is the same as used in   Markov-Chain Monte-Carlo is
used to fit source parameters and additional nuisance parameters (see
below) jointly, and then marginalise out all uninteresting parameters,
leaving posterior distributions for any chosen parameter.

In addition to the source parameters, the fitting procedure includes
three nuisance parameters, namely the brightness normalisation and the
start and end points on the path.  The latter two are equivalent to
considering the location of the caustic with respect to the starting
point and the velocity component of the source transverse to the
caustic (that is, $v_{\!\perp}$).  The orientation of the source track
is another nuisance parameter, which we do not include it, because for
a pure fold, changing the orientation of the source track is
equivalent to changing $v_{\!\perp}$ and the orientation of the source
(see Figure~\ref{fig:infinite_fold}), all parameters that are
included.  Since the caustic is not a perfect fold, disregarding the
orientation of the source track can be expected to introduce some
systematic error.

Figures \ref{fig:cc_forward}--\ref{fig:gc_forward} show the numerical
results.  Each of these figures shows the mock data (the same in all
three cases) along with best fits and the fitted values and
uncertainties on $r_{1/2}$ and (if applicable) $R_n/R_p$.

Figure \ref{fig:cc_forward} shows a test of the fitting procedure.
The computer compares the mock data to different templates
lightcurves.  All the templates come from taking a crescent source
along the path CB in Figure~\ref{fig:magnification_map}, but with
varied crescent parameters and varied start and finish points.  As the
figure shows, the lightcurve can be fitted to within the noise, and
the parameters $r_{1/2}$ and $R_n/R_p$ can be recovered.

Figure \ref{fig:cc_backward} attempts a somewhat more realistic
situation.  The mock lightcurve is the same as before, but the
templates come from taking a crescent source along the path AB in
Figure~\ref{fig:magnification_map}.  That is, the mock data and the
templates come from crossings of different caustics.  This mimics the
unavoidable systematic error of not knowing the caustic exactly.  The
fit is slightly degraded, but still acceptable.  The parameters
$r_{1/2}$ and $R_n/R_p$ are still recovered.  The parameters $a/R_p$
and $b/R_p$ were not correctly recovered, which is understandable from
the track change.

Figure \ref{fig:gc_forward} shows the result of attempting to fit
templates from a Gaussian source to the same mock data.  There is a
small but statistically very significant misfit because a Gaussian
source cannot reproduce the part of the lightcurve with the slight
flattening and then a second sub-peak
(cf.~Figure~\ref{fig:char_points}) as the dark part of the crescent
crosses the caustic.  A hint of such a feature is present in recent
caustic-crossing observations of the well-known system Q2237+0305
\citep[see Figure~3 in][]{2015ApJ...814L..26M} though interpretation
as a crescent source is speculation at present.
